---
title: "Citations of Retracted ICPSR Papers"
output: pdf_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(parallel)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE)
# setwd(...)

# load data
merged <- read_csv("merged.csv")
refs <- read_csv("citations/refs_merged.csv")
```

Number of citations:

```{r}
# add file id column to merged - TODO: move to dc/preprocessing file?
merged <- merged |> mutate(file.id = NA, .before = Title)

merged$file.id[which(merged$Title == 
  "Worlds Apart? The Reception of Genetically Modified Foods in Europe and the U.S."
  )] = "WorldsApart"
merged$file.id[which(merged$Title == 
  "Nonmetropolitan sex-role ideologies: A longitudinal study"
  )] = "NonmetroRole"
merged$file.id[which(merged$Title ==
  "School social bonds, school climate, and school misbehavior: A multilevel analysis"
  )] = "SchoolSocialBonds"
merged$file.id[which(merged$Title ==
  "Suicide after natural disasters"
  )] = "SuicideNatDisasters"
merged$file.id[which(merged$Title ==
  "Electronic cigarette use and myocardial infarction among adults in the US Population Assessment of Tobacco and Health"
  )] = "ECigUse"
merged$file.id[which(merged$Title ==
  "The verdict is in: How did they decide? Using drivers' self-reported data to understand officers' decision making during traffic stop encounters"
  )] = "HowDidTheyDecide"
merged$file.id[which(merged$Title ==
  "Chronic adolescent marijuana use as a risk factor for physical and mental health problems in young adult men"
  )] = "MarijuanaUse"
merged$file.id[which(merged$Title ==
  "Tuning in, not turning out: Evaluating the impact of ethnic television on political participation"
  )] = "TurningIn"
merged$file.id[which(merged$Title ==
  "Implications of new hypertension guidelines in the United States"
  )] = "HypertensionGuidelines"

# add retraction dates to refs
refs <- refs |> mutate(ret.date = NA, .after = year)

for (r in 1:nrow(refs)) {
  ix = which(refs$filepath[r] == merged$file.id)
  refs$ret.date[r] = merged$RetractionDate[ix]
}

# add retraction years to refs
refs <- refs |> mutate(ret.year = NA, .after = ret.date)

for (r in 1:nrow(refs)) {
  ix = which(refs$filepath[r] == merged$file.id)
  refs$ret.year[r] = merged$RetractionYear[ix]
}

# number of citations per paper
refs |> group_by(filepath) |> count()

# number of citations per paper by year
refsbyyear <- refs |> group_by(filepath, year, ret.year) |>
  summarise(num_citations = n())
(head(refsbyyear, 20))
```


Plots of citations by year. Retraction Year in red.
```{r}
# plot citations by year
refsbyyear |> 
  group_by(filepath) |>
  ggplot(aes(x = year, y = num_citations)) +
    geom_line() +
    geom_vline(aes(xintercept = ret.year), colour = "red") +
    xlim(1990, 2030) +
    ylim(0, 50) +
    facet_wrap(~filepath)

```

Summary table - number of citations before and after retraction year:

```{r}
refsbyyear |> group_by(filepath, year <= ret.year) |>
  count()
```

Flag whether a retracted article was cited the same year it was retracted:

```{r}
# create flag
refs <- refs |> mutate(ref.same.year = (year == ret.year))

# number citing papers retracted during the publication year
refs |> group_by(ref.same.year) |> summarise(n())

# number with available publication date
refs |> group_by(ref.same.year, !is.na(pub.date)) |> summarise(n())
```

Pulling citation data manually for retractions for which `ret.same.year == TRUE`:

```{r}
# setwd(...)

# refs |> filter(ref.same.year == TRUE) |>
#   select(title, authors.list, year, pub.date, doi) |>
#   write.csv(file = "refs_same_year.csv")

# manually pull and add dates to sheet, export to CSV

# Searched on WOS and Scopus. Could typically find data on WOS 
# directly, on Scopus, viewed article page on Publisher
# website or on Proquest/SagePub/EBSCOHost through UM library
```

Manually downloaded data:

```{r}
# read edited file into new df
refs_manual <- read.csv("refs_same_year_final.csv")
```


Reformat pub.date

```{r}
month.names <- c(
  'JAN' = 1,
  'FEB' = 2,
  'MAR' = 3,
  'APR' = 4,
  'MAY' = 5,
  'JUN' = 6,
  'JUL' = 7,
  'AUG' = 8,
  'SEP' = 9,
  'OCT' = 10,
  'NOV' = 11,
  'DEC' = 12
)

get.date <- function(datestr) {
  # split into vector of (MONTH, [DAY], YEAR)
  datevec = str_split(datestr, pattern = '\\s') |> unlist()
  
  # convert year, month to numeric
  # check for date, if yes, convert to numeric
  year = datevec[length(datevec)]
  month = month.names[datevec[1]] |> unname()
  day = ifelse(length(datevec) > 2, datevec[2], NA)
  
  return(c(year, month, day))
}

refs$pub.year = numeric(nrow(refs))
refs$pub.month = numeric(nrow(refs))
refs$pub.day = numeric(nrow(refs))

for (r in 1:nrow(refs)) {
  # if pub.date data is provided, use get.date function
  if (!is.na(refs$pub.date[r])) {
    datevec = get.date(refs$pub.date[r])
    refs$pub.year[r] = datevec[1]
    refs$pub.month[r] = datevec[2]
    refs$pub.day[r] = datevec[3]
  }
  # if pub.date is not provided, use manually pulled data if available
  else {
    # check if doi is available
    if (!is.na(refs$doi[r])) {
      ix = which(refs$doi[r] == refs_manual$doi)
    }
    # else compare titles, publication years for match
    else {
      ix = which((refs$title[r] == refs_manual$title) &
                   (refs$year[r] == refs_manual$year))
    }
    
    # check whether replacement exists
    if (length(ix) == 0) {
      # if not, insert NAs
      refs$pub.year[r] = NA
      refs$pub.month[r] = NA
      refs$pub.day[r] = NA
    }
    # if so, replace
    else {
      refs$pub.year[r] = refs_manual$pub.year.manual[ix]
      refs$pub.month[r] = refs_manual$pub.month.manual[ix]
      refs$pub.day[r] = refs_manual$pub.date.manual[ix]
    }
  }
}
```

Flag if month AND year of ref publication are same as retraction month and year

```{r}
# add retraction months to refs
refs <- refs |> mutate(
  ret.month = mdy_hm(ret.date) %>% month(),
  .after = ret.year
)

# add retraction date specifically to refs
refs <- refs |> mutate(
  ret.day = mdy_hm(ret.date) %>% mday(),
  .after = ret.month
)

# number citing papers retracted during the publication month
refs |> group_by((year == ret.year) & (pub.month == ret.month)) |> summarise(n())
```

More granular before/after retraction counts - only 3 refs published in retraction year for which no month could be obtained (1 publication citing "Suicide after natural disasters...", 2 citing "Worlds Apart? ...")

```{r}
refs |> group_by(filepath, ((year < ret.year) | ((year == ret.year) & (pub.month < ret.month)))) |> 
  count()
```

Normalize retraction year, line graph citations against normalized retraction year:

```{r}
refs <- refs |> mutate(pub.year.norm = year - ret.year)


refs |> group_by(filepath, pub.year.norm) |>
  count() |>
  ggplot(aes(x = pub.year.norm, y = n, color = filepath)) +
    geom_line(linewidth = 0.75) +
    geom_vline(aes(xintercept = 0), linewidth = 1, linetype = "dashed",
               color = "black", alpha = 0.3) +
    xlab("Years pre- or post-retraction") +
    ylab("Total annual citations") +
    theme(legend.title = "Paper ID") +
    theme_linedraw()
```

Bubble plot:

```{r}
refs |> group_by(filepath, year, pub.year.norm) |>
  count() |>
  ggplot(aes(x = year, y = pub.year.norm, size = n, colour = filepath)) +
    geom_point(alpha = 0.6) +
    labs(x = "Year", y = "Years pre- or post-retraction",
         size = "Number of citations", colour = "Paper ID")
```

